---
title: "The memory wall, measured: 18 years of ML accelerators"
description: "Peak compute on AI chips doubles almost twice as fast as memory bandwidth, and tensor cores cause the whole gap. A trend analysis in R, and what it means for LLM inference."
author: "Evan Pai"
date: 2026-09-21
categories: [R, hardware, HPC, data analysis]
execute:
  echo: true
---

Every GPU launch comes with a headline FLOP/s number, and every generation that
number jumps. When I was tuning LLM inference on H100s, though, the question
that kept coming up was how fast weights could be pulled out of memory, not how
fast they could be multiplied. This post checks that intuition against data: across every
ML accelerator Epoch AI has catalogued, how fast has peak compute grown, how fast
has memory bandwidth grown, and how wide is the gap now?

**Data.** [Data on Machine Learning Hardware](https://epoch.ai/data/machine-learning-hardware)
by Epoch AI, licensed under
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Citation: Epoch AI,
"Data on Machine Learning Hardware". Published online at epoch.ai. Retrieved from
<https://epoch.ai/data/machine-learning-hardware> [online resource]. Accessed
23 Sep 2026. The dataset is updated continuously, so I committed the snapshot
downloaded on 23 September 2026 (`ml_hardware.csv`, about 100 KB) next to this post.
That way the numbers below do not change between builds.

## Loading the data

```{r}
#| label: setup
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(knitr)
```

```{r}
#| label: load
hardware_raw <- read_csv("ml_hardware.csv", show_col_types = FALSE)
dim(hardware_raw)
```

Each row is one chip: 176 accelerators, with 39 columns of specs. Most of those
columns are sparsely filled. Vendors publish different numbers, and precision
formats change between generations. So the first job is to decide which columns
to trust and what to do when one is missing.

## Cleaning: one comparable "peak compute" number per chip

Chips before 2017 have no tensor cores, and many recent chips only publish
tensor-core numbers. I keep three columns: tensor FP16/BF16, plain FP16 and
FP32. The headline "peak" is the **largest** of the three. Taking the largest
rather than preferring 16-bit matters. Consumer Pascal cards from 2016 run FP16
at 1/64 of their FP32 rate, so preferring 16-bit would record those chips as
roughly 60× slower than they really are. The `precision` column records which
format supplied the peak, so the mix stays visible.

```{r}
#| label: clean
hardware <- hardware_raw |>
  transmute(
    name         = `Hardware name`,
    manufacturer = Manufacturer,
    type         = Type,
    release      = as.Date(`Release date`),
    tensor16     = `Tensor-FP16/BF16 performance (FLOP/s)`,
    fp16         = `FP16 (half precision) performance (FLOP/s)`,
    fp32         = `FP32 (single precision) performance (FLOP/s)`,
    bandwidth    = `Memory bandwidth (byte/s)`,
    memory       = `Memory (bytes)`,
    price        = `Release price (USD)`
  ) |>
  mutate(
    peak = pmax(tensor16, fp16, fp32, na.rm = TRUE),
    precision = case_when(
      peak == tensor16 ~ "Tensor FP16/BF16",
      peak == fp16     ~ "FP16",
      peak == fp32     ~ "FP32",
      .default         = NA_character_
    ),
    year = as.numeric(format(release, "%Y")) +
      (as.numeric(format(release, "%j")) - 1) / 365.25
  ) |>
  filter(!is.na(release), peak > 0 | bandwidth > 0)

hardware |>
  summarise(
    chips = n(),
    first_release = min(release),
    last_release = max(release),
    with_peak = sum(!is.na(peak) & peak > 0),
    with_bandwidth = sum(!is.na(bandwidth)),
    with_both = sum(!is.na(peak) & peak > 0 & !is.na(bandwidth))
  ) |>
  kable(caption = "Coverage of the cleaned hardware table.")
```

## Compute and bandwidth over time

```{r}
#| label: fig-compute-vs-bandwidth
#| fig-cap: "Peak compute (FLOP/s) and memory bandwidth (byte/s) of ML accelerators by release date, on a log scale. Lines are least-squares fits of log2(value) on release year."
#| fig-width: 8
#| fig-height: 4.8
trend_data <- hardware |>
  select(name, year, `Peak compute (FLOP/s)` = peak,
         `Memory bandwidth (byte/s)` = bandwidth) |>
  pivot_longer(-c(name, year), names_to = "metric", values_to = "value") |>
  filter(!is.na(value), value > 0)

ggplot(trend_data, aes(year, value, colour = metric)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, linewidth = 0.9) +
  scale_y_log10(labels = label_log()) +
  scale_colour_manual(values = c("#1b6ca8", "#d1495b")) +
  labs(x = "Release year", y = NULL, colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

The two clouds in @fig-compute-vs-bandwidth start about three orders of
magnitude apart and drift further apart every year. The slopes can be turned into
a number that is easier to reason about: the **doubling time**. If
`log2(value) = a + b * year`, then the value doubles every `1 / b` years. I fit
one regression per metric and turn the 95% confidence interval on `b` into an
interval on the doubling time.

```{r}
#| label: doubling-times
doubling <- function(df, column) {
  d <- df |> filter(!is.na(.data[[column]]), .data[[column]] > 0)
  fit <- lm(log2(d[[column]]) ~ d$year)
  slope <- coef(fit)[2]
  ci <- confint(fit)[2, ]
  tibble(
    n = nrow(d),
    `growth per year` = sprintf("%.2fx", 2^slope),
    `doubling time (years)` = sprintf("%.2f", 1 / slope),
    `95% CI` = sprintf("%.2f to %.2f", 1 / ci[2], 1 / ci[1])
  )
}

doubling_table <- bind_rows(
  `Peak compute, any format`          = doubling(hardware, "peak"),
  `FP32 compute only`                 = doubling(hardware, "fp32"),
  `Memory bandwidth`                  = doubling(hardware, "bandwidth"),
  `Memory capacity`                   = doubling(hardware, "memory"),
  .id = "metric"
)
kable(doubling_table, caption = "Doubling times from log-linear fits over all chips with a release date.")
```

Peak compute doubles about every year and a half. Memory bandwidth and capacity
take close to three years. The FP32-only row was meant as a robustness check,
and it changed my reading of the result. **FP32 compute has grown at almost
exactly the same rate as bandwidth.** The confidence intervals overlap almost
completely. On the chips' traditional arithmetic path, the memory wall does not
exist in this data. The entire gap comes from the tensor cores and lower-precision
formats that vendors added from 2017 onward. So the memory wall is not a law of
silicon. It is a consequence of the industry spending its transistor budget on
dense low-precision matrix units.

## What the gap means: the roofline ridge point

The roofline model gives this gap a concrete meaning. A kernel with
*arithmetic intensity* $I$ (FLOPs performed per byte moved from memory) can run
at most at $\min(\text{peak},\ I \times \text{bandwidth})$. The crossover
$I^\ast = \text{peak} / \text{bandwidth}$ is the **ridge point**. Kernels below
it are memory-bound and kernels above it are compute-bound. If compute grows
faster than bandwidth, the ridge point moves right, and more real workloads end
up memory-bound. @fig-ridge-point and the table below show exactly that
happening, in step with the spread of tensor cores.

```{r}
#| label: fig-ridge-point
#| fig-cap: "Ridge point (peak FLOP/s divided by memory bandwidth) of each accelerator. A kernel needs at least this many FLOPs per byte of memory traffic to be compute-bound. Labelled points are data-centre flagships."
#| fig-width: 8
#| fig-height: 4.8
flagships <- c(
  "NVIDIA Tesla V100 SXM2 16 GB", "NVIDIA A100 SXM4 80 GB",
  "NVIDIA H100 SXM5 80GB", "NVIDIA B200", "NVIDIA GB300 (Blackwell Ultra)",
  "AMD Instinct MI300X", "AMD Instinct MI355X",
  "Google TPU v3", "Google TPU v4", "Google TPU v7 Ironwood"
)

ridge <- hardware |>
  filter(!is.na(peak), peak > 0, !is.na(bandwidth)) |>
  mutate(ridge = peak / bandwidth, flagship = name %in% flagships)
labelled <- c("NVIDIA Tesla V100 SXM2 16 GB", "NVIDIA H100 SXM5 80GB",
              "NVIDIA GB300 (Blackwell Ultra)")

ggplot(ridge, aes(year, ridge)) +
  geom_point(aes(colour = precision), alpha = 0.7, size = 2) +
  geom_text(
    data = filter(ridge, name %in% labelled),
    aes(label = sub(" \\(.*\\)", "", sub("NVIDIA (Tesla )?", "", name))),
    size = 3, vjust = -1, hjust = 0.8
  ) +
  scale_y_log10() +
  scale_colour_manual(values = c("#8a8a8a", "#e09f3e", "#1b6ca8")) +
  labs(x = "Release year", y = "FLOPs per byte (log scale)", colour = "Peak measured as") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r}
#| label: ridge-by-era
ridge |>
  mutate(era = cut(year, c(2008, 2016, 2020, 2023, 2027), right = FALSE,
                   labels = c("2008–2015", "2016–2019", "2020–2022", "2023–2026"))) |>
  group_by(era) |>
  summarise(
    chips = n(),
    `median ridge (FLOP/byte)` = round(median(ridge)),
    `share with tensor-core peak` = percent(mean(precision == "Tensor FP16/BF16"))
  ) |>
  kable(caption = "Median ridge point by release era.")
```

## Consequence for LLM inference

The ridge point explains something every LLM serving engineer runs into. When
you generate one token at batch size 1, every weight is read once and used for
about two FLOPs (one multiply, one add). With 16-bit weights that is 2 bytes per
parameter, so the arithmetic intensity is roughly **1 FLOP per byte**. The
flagships above have ridge points in the hundreds. Single-stream decoding
therefore uses well under 1% of the chip's peak compute, and memory bandwidth
sets the speed limit:

$$
\text{tokens/s} \;\le\; \frac{\text{bandwidth}}{2 \times \text{parameters}}.
$$

```{r}
#| label: decode-bound
params <- 70e9  # a 70B-parameter model in 16-bit weights

ridge |>
  filter(flagship) |>
  arrange(release) |>
  transmute(
    chip = name,
    released = format(release, "%Y-%m"),
    `peak (TFLOP/s)` = round(peak / 1e12),
    `bandwidth (TB/s)` = round(bandwidth / 1e12, 2),
    `ridge (FLOP/byte)` = round(ridge),
    `70B decode ceiling (tok/s, batch 1)` = round(bandwidth / (2 * params), 1),
    `peak compute used at batch 1` = percent(1 / ridge, accuracy = 0.01)
  ) |>
  kable(caption = "Bandwidth-bound decode ceiling for a 70B model on one chip, ignoring KV-cache traffic and assuming the weights fit.")
```

This is why serving systems like vLLM work so hard on **batching**. With a
batch of $B$ sequences, each weight read from memory is used $B$ times, so
arithmetic intensity grows roughly linearly in $B$. The batch can climb toward
the ridge point before compute becomes the limit. Continuous batching was one
of the main levers when I tuned a LLaMA 3.1 70B chatbot on H100s for
throughput, and this table shows why.

## What I found

- Across 18 years of accelerators, **peak compute doubled every ~1.6 years
  against ~2.8 years for memory bandwidth**.
- **The gap comes from tensor cores.** FP32 compute kept pace with bandwidth. The
  divergence is the low-precision matrix hardware added since 2017.
- The **median ridge point rose from 17 to over 200 FLOPs per byte**, more than
  an order of magnitude. A kernel now needs far more reuse of each byte before it
  can use the chip it runs on.
- For LLM decoding at small batch sizes, that makes **bandwidth, not FLOP/s,
  the number to read on a spec sheet**, and batching, KV-cache layout and
  quantisation the levers that matter.

**Caveats.** These are vendor peak specs, not measured throughput. Each chip
counts once however widely it was deployed. The log-linear fit treats a
consumer card and a TPU pod chip as the same kind of point. A deployment-weighted
analysis would be a natural next step.
