---
title: "How much of the GPU did they actually use? R and Python in one post"
description: "Joining Epoch AI hardware specs (cleaned in R) with model records (processed in Python) to estimate how much of the GPU real training runs used, and catching a circular estimate before it became a finding. Data frames pass between the languages with reticulate."
author: "Evan Pai"
date: 2026-09-23
categories: [R, Python, reticulate, HPC]
engine: knitr
---

My [R post](../gpu-hardware-trends/index.qmd) looked at what accelerators *can*
do. My [Python post](../ai-compute-scaling/index.qmd) looked at how much compute
models *used*. This post joins the two to ask what every cluster engineer asks
about a training run: **what fraction of the hardware's peak did it achieve?**

That number, often called model FLOPs utilisation (MFU), can be estimated from
public records:

$$
\text{utilisation} \approx \frac{\text{training compute (FLOP)}}
{\text{chips} \times \text{peak FLOP/s per chip} \times \text{training time (s)}}.
$$

The work is split between the two languages. **R** cleans the hardware specs,
using the same approach as the R post. **Python** reads the model records,
joins them and computes utilisation. **R** then plots the result. Objects cross
between them through [reticulate](https://rstudio.github.io/reticulate/): Python
reads R objects as `r.name`, and R reads Python objects as `py$name`.

**Data.** Two Epoch AI datasets, both licensed under
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/):
[Data on Machine Learning Hardware](https://epoch.ai/data/machine-learning-hardware)
and [Data on AI Models](https://epoch.ai/data/ai-models) (Epoch AI, published
online at epoch.ai, accessed 23 Sep 2026). This post reuses the snapshots
committed next to the two earlier posts, so no data is downloaded at build
time.

## Step 1 (R): clean the hardware table

```{r}
#| label: setup
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(reticulate)
```

```{r}
#| label: r-hardware
hardware <- read_csv("../gpu-hardware-trends/ml_hardware.csv", show_col_types = FALSE) |>
  transmute(
    hardware = `Hardware name`,
    peak_flops = pmax(`Tensor-FP16/BF16 performance (FLOP/s)`,
                      `FP16 (half precision) performance (FLOP/s)`,
                      `FP32 (single precision) performance (FLOP/s)`, na.rm = TRUE)
  ) |>
  filter(!is.na(peak_flops), peak_flops > 0)

nrow(hardware)
```

This is a plain R `data.frame`. The next chunk is Python and reads it directly
as `r.hardware`.

## Step 2 (Python): join model records and compute utilisation

```{python}
#| label: py-environment
import sys
print(sys.executable)  # should be this project's .venv
```

```{python}
#| label: py-join
import numpy as np
import pandas as pd

hardware = r.hardware  # the R data frame, now a pandas DataFrame
print(type(hardware).__name__, hardware.shape)

models = pd.read_csv("../ai-compute-scaling/notable_ai_models.csv")
runs = models.dropna(subset=["Training compute (FLOP)", "Training hardware",
                             "Hardware quantity", "Training time (hours)"]).copy()
runs["year"] = pd.to_datetime(runs["Publication date"]).dt.year
matched = runs.merge(hardware, left_on="Training hardware", right_on="hardware", how="inner")
print(f"{len(runs)} runs with compute, hardware, chip count and time; "
      f"{len(matched)} match a chip in the hardware table")

matched["utilisation"] = matched["Training compute (FLOP)"] / (
    matched["Hardware quantity"] * matched["peak_flops"] * matched["Training time (hours)"] * 3600
)
matched["plausible"] = matched["utilisation"].between(0.01, 1.0)
```

### A circularity check before trusting the answer

A first plot of these numbers showed a suspicious pile-up just above 30%. Epoch's
`Training compute notes` column explains it. When a lab does not report compute,
Epoch often **estimates it from the hardware**, as chips × peak × time × an
*assumed* utilisation, usually 30–40%. For those rows, the formula above just
returns Epoch's assumption. It is circular.

So before measuring anything, I tag each run by how its compute was estimated.
If the notes mention an assumed utilisation or an "estimate from hardware", the
run is tagged *hardware-based*. The others are *independent*: reported by the
developer, or counted from parameters and tokens. Only the independent runs say
anything about real utilisation. The tagging is a keyword heuristic, so it is
approximate. The pattern is written out below so it can be checked.

```{python}
#| label: py-circularity
assumed_pattern = (
    r"assum\w*[^.]{0,80}utili[sz]\w*"        # "30% utilization assumed", "assume 0.4 utilization"
    r"|utili[sz]\w*[^.]{0,40}assum\w*"
    r"|\d+\s*% (?:hardware )?utili[sz]\w*"   # "30% utilization"
    r"|0\.\d+ ?(?:hardware )?utili[sz]\w*"   # "0.3 utilization"
    r"|utili[sz]ation rate of"
    r"|hardware details|estimate from hardware"
)
notes = matched["Training compute notes"].fillna("")
matched["estimate"] = np.where(
    notes.str.contains(assumed_pattern, case=False, regex=True),
    "Hardware-based (utilisation assumed)",
    "Independent (reported or counted)",
)

utilisation = matched[["Model", "year", "hardware", "Hardware quantity", "utilisation",
                       "plausible", "estimate"]].rename(columns={"Hardware quantity": "chips"})
utilisation.groupby("estimate").agg(
    runs=("utilisation", "size"),
    flagged_implausible=("plausible", lambda p: int((~p).sum())),
)
```

Runs outside 1–100% are flagged, not dropped. An estimate above 100% is
physically impossible, so at least one input must be off. Common causes are a
run trained in FP8 measured against a 16-bit peak, a chip count that includes
idle spares, or a training time that leaves out restarts.

## Step 3 (R): plot what Python computed

Back in R, `py$utilisation` is the pandas DataFrame from Step 2, converted to an
R `data.frame`.

```{r}
#| label: fig-utilisation
#| fig-cap: "Estimated hardware utilisation of training runs, computed in Python and plotted in R, split by how Epoch estimated each run's compute. Only plausible runs (1–100%) are shown. The shaded band is 28–42%, the range of utilisation values Epoch typically assumes."
util <- py$utilisation

util |>
  filter(plausible) |>
  ggplot(aes(year, utilisation)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.28, ymax = 0.42,
           fill = "#e09f3e", alpha = 0.18) +
  geom_jitter(aes(size = chips), width = 0.25, height = 0, colour = "#1b6ca8", alpha = 0.6) +
  facet_wrap(~estimate, ncol = 1) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_size_area(max_size = 6, labels = scales::comma) +
  labs(x = "Publication year", y = "Estimated utilisation", size = "Chips") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: r-summary
util |>
  filter(plausible) |>
  group_by(estimate) |>
  summarise(
    runs = n(),
    median = scales::percent(median(utilisation), accuracy = 1),
    `middle 50%` = paste(scales::percent(quantile(utilisation, c(0.25, 0.75)), accuracy = 1),
                         collapse = " to "),
    `inside the 28–42% band` = scales::percent(mean(between(utilisation, 0.28, 0.42)), accuracy = 1),
    `median chips` = median(chips)
  ) |>
  kable(caption = "Utilisation of plausible runs, by how the run's compute was estimated.")
```

## What I found

- **Half the data would have answered with its own assumption.** More than
  half of the runs that can be joined have compute estimated from the hardware,
  and most of them land inside the assumed 28–42% band. A naive analysis would
  have "discovered" that training runs achieve about 30% utilisation, which is
  exactly what was put in.
- **The independent runs are more informative, and they are wider.** Their
  median is also close to 30%, so Epoch's default is a reasonable prior. The
  spread is much larger, though. Real runs range from well under 20% to over 50%,
  and that spread is where a cluster team's engineering work shows up.
- **Physical bounds are a cheap data check.** Requiring 1–100% utilisation caught
  inconsistent records that a purely statistical filter would have kept.

**How the languages split the work.** R's `pmax(..., na.rm = TRUE)` made the
"best available precision" rule a one-liner. pandas handled the join, the regex
tagging and the arithmetic. ggplot2 drew facets from a data frame created in
Python. Nothing was written to disk between steps. The objects passed in memory
through `r.hardware` and `py$utilisation`.
